---
title: "Reproducible Research: Peer Assessment 1"
author: "Kum Buh Paul"
date:  "Sept 2017"
output: html_document
---

This is the first project for Reproducible Research course in the datascience specialization. This project answer a series of questions related to data collected from [FitBit](https://en.wikipedia.org/wiki/Fitbit)

##Data
The data for the analysis was downloaded from course web site.

  Dataset: [Activity monitoring data](https://arabia.starzplay.com/en/landing?utm_source=affiliate&utm_medium=desktop&utm_campaign=propc&utm_content=1039081&data2=359583771131)
  
The dataset includes 3 variables:

- steps: Number of steps taking in a 5-minute interval(missing values are coded as NA)
- date: The date on which the measurement was taken in YYY-MM-DD format
- interval: Identifier for the 5-minute interval in which measurement was taken.

The dataset is stored in a comma-separated-value(CSV) file and there are a total of 17,558 observations



The data used for this analysis is contained in a file called activity.csv. This code could be written to download incase the file isn't present.

## Downloading Data
```{r echo=TRUE}
if(!file.exists("activity.zip")) {
  #download the data
  fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(fileUrl, destfile = "activity.zip")
  
  # unzip file
  unzip("AMD.zip")
}
```

## Loading and preprocessing the data


```{r}
#load data
activity <- read.csv("activity.csv")
```

## Tidying Data

Change variable date into dateformat using lubridate package
```{r}
library(lubridate)
activity$date <- ymd(activity$date)
```

Check the activity dataset with str() and head()


## What is mean total number of steps taken per day?

Using the dplyr library, we will group the activity dataset by date, then summarise it with the mean function while ignoring the missing values.
```{r}
library(dplyr)
steps_day <- summarise(group_by(activity, date), "steps" = sum(steps, na.rm = T))
print(steps_day)

# create a histogram

hist(steps_day$steps, xlab = "number of steps per day", main = "Total steps taken each day", col = "blue")

#mean and median of the total steps
meanNumber <- mean(steps_day$steps); meanNumber
medianNumber <- median(steps_day$steps); medianNumber


```

## What is the average daily activity pattern?
To answer this question, a time series plot will be produced for the average total number of steps taken for each interval across all days.

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?.



###Methodology and Result


1. Calculate the average number of steps taken in each 5-minute interval per day using dplyr and group by interval:

```{r}
# Average number of steps taken
averageSteps <- summarise(group_by(activity, interval), "steps" = mean(steps, na.rm = T))
print(averageSteps)

```

Use base ploting system for making the time series of the 5-minute interval and average steps taken:

```{r}

with(averageSteps, plot(interval, steps, type = "l", xlab = "Interval", ylab = "Number of Steps", main="Average Number of Steps per Day by Interval"))
```

2. Use which.max() to find out the maximum steps, on average, across all the days:

```{r}
maxi <- averageSteps[which.max(averageSteps$steps), ]
print(maxi)
```

```{r echo = FALSE}
maxInterval <- maxi$interval
maxSteps <- maxi$steps
```
The interval `r maxInterval ` has, on average, the highest count of steps, with 206 steps.


## Imputing missing values

1. Calculate and report missing values
```{r}
missingValue <- sum(is.na(activity$steps))

```


```{r echo=FALSE}
missingValue
```
There are a total of `r missingValue` missing values for the step variable in activity data

2. Let's take the approach to fill in a missing NA with the average number of steps in the same 5-min interval.
3. Create a new dataset as the original and use transform, ifelse and match function for filling in the missing values with the average number of steps per 5-minute interval:

```{r}
imputed_data <- transform(activity , steps = ifelse(is.na(activity$steps), averageSteps$steps[match(activity$interval, averageSteps$interval)], activity$steps))
print(head(imputed_data))

# Check that there are no missing values
sum(is.na(imputed_data$steps))
```

4. Calculate the number of steps taken in each 5-minute interval per day using dplyr and group by interval. Use base plot system for making the histogram

```{r}
full_steps <- summarise(group_by(imputed_data, date), steps = sum(steps))
print(full_steps)

hist(full_steps$steps, xlab = "number of steps per day", main = "Total steps taken each day", col = "violet")

# Mean and Median

mean_full <- mean(full_steps$steps, na.rm = TRUE)
mean_full
median_full <- median(full_steps$steps, na.rm = TRUE)
median_full
```
The impact of imputing missing data with the average number of steps in the same 5-min interval is that both the mean and the median are equal to the same value: 10766.


## Are there differences in activity patterns between weekdays and weekends?

Use dplyr and mutate to create a new column, weektype, and apply whether the day is weekend or weekday:
```{r}
library(lubridate)

imputed_data <- mutate(imputed_data, weektype = ifelse(weekdays(imputed_data$date) == "Saturday" | weekdays(imputed_data$date) == "Sunday", "weekend", "weekday"))
imputed_data$weektype <- as.factor(imputed_data$weektype)
head(imputed_data)
```

2. Calculate the average steps in the 5-minute interval and use ggplot2 package for making the time series of the 5-minute interval for weekday and weekend, and compare the average steps:

```{r}
library(ggplot2)
interval_full <- imputed_data %>%group_by(interval, weektype) %>% summarise(steps = mean(steps))

s <- ggplot(interval_full, aes(interval, steps, color = weektype)) +
  geom_line() +
  facet_wrap(~weektype, ncol = 1, nrow = 2)
print(s)
```